Flattering Advice: Avoiding Disappointment as a Driver of Gender Discrimination

### 2024-09-30 manuscript/manuscript.Rmd
Method part

<!-- David: Just commenting on what I tried to do here... a bit of a transition -->

<!-- for why this study follows from the last. -->

So far, we have shown that displaying expectations can induce gender
differences in advice-giving. Specifically, advisers were more likely to
recommend competing against a group of High Performers to men who had
higher expectations about their performance than did women. We now test
directly whether this is driven by interpersonal concerns. Specifically,
we examine whether more flattering advice is the result of a desire to
be liked and whether advisees indeed do like advisers more when they
give flattering advice.

We conducted a three-stage experiment in which we experimentally
manipulated the incentives for advicers. We began by inviting a group of
advisees to upload photos of themselves ("selfies") and informed them
that they would be rated on their attractiveness. We grouped them with
nine other participants of the same gender and recruited participants of
the opposing gender to rank them from most to least attractive and to
provide advice. Specifically, we asked them to advise the participant
they ranked as the 7th most attractive (4th least attractive) on what
rank they should bet they were ranked by a larger group of raters.
Advisers were randomly assigned to two treatments, receiving a bonus
payment either if the advisee guessed their rank accurately or if the
advisee evaluated the adviser as likeable as measured by an
unincentivized scale response. We hypothesize that advisers who want to
be liked by advisees will recommend that they bet on a lower rank, i.e.
that they are more attractive.

@@

After submitting their ratings, they saw the photo of the participant
they had ranked as the 7th most attractive (i.e., the 4th least
attractive). They were reminded of the rank they had given to that perso
and informed that this participant would be invited back and could earn
a \$1 bonus if they guessed their rank correctly. The rank was
determined by the aggregate ratings of all participants who had ranked
this group. Their task was to give advice to this participant about the
rank they should bet on based on having observed all ten selfies and
their own assessment. We randomly assigned participants to one of two
incentivization schemes. In the "Accuracy" treatment, they received a
bonus identical to the advisee: \$1 if they guessed their rank
correctly. In the "Likeability" treatment, we informed them that they
would be rated by the advisee on a 5-point Likert scale for how likeable
they thought they were. Each point on the scale would translate to a
bonus of 20 cents. They then selected a rank that they would recommend
the advisee to bet on.

@@

<!-- Amanda: How about this -->

<!-- For warmth it looks interesting, but I think the plot is hard -->

<!-- to read. Let's hold off on this for now. -->

@@

<!-- The Figure is not super clear. How about we try a series of box plots, -->

<!-- with the x-axis as the advised rank? Standard errors will be huge at the -->

<!-- extremes. -->

<!-- Also, we need to do something relative to the advisees' expectations. -->

<!-- I wonder what would be interesting here? Maybe averages for three groups: -->

<!-- more attractive than the participants' own guess, equally attractive, and less -->

<!-- attractive? -->

<!-- Amanda: How about the following picture? -->

@@

library(dplyr)
library(ggplot2)

study2_time3 <- study2_time3 %>% dplyr::mutate(Guess = as.numeric(Guess), Advice = as.numeric(Advice)) %>%
  dplyr::mutate(HorL = ifelse(Guess == Advice,0, ifelse(Guess > Advice, 1,-1))) %>%
  dplyr::mutate(HorL = as.factor(HorL)) %>% 
  dplyr::mutate(bonus = ifelse(bet == `Final Rank`,1,0))

ratings <- study2_time3 %>%
  group_by(HorL) %>%
  summarise(like = mean(likable), warmth = mean(warmAvg))

rating_table <- matrix(c(ratings$like,ratings$warmth), nrow = 2, byrow = TRUE, dimnames = list( c("Likable",'Warm'),c("Pessimistic", "Accurate", "Flattering")))

@@

ggplot(rating_df, aes(x = Advice, y = Freq, fill = Condition)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("#2A629A",'#FF7F3E')) +
  labs(x = "Advice", y = "Rating") +
  ggtitle('Response to Advice') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),          
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20)) +
  geom_text(aes(label = paste0(Freq), y = 0.5), position = position_dodge(width = 0.9), size = 5, color = 'white', fontface = "bold")

@@

We were not powered to do a comparison across the two experimental
groups and did not preregister such a difference. Indeed, we find no
difference in likeability and warmth across the two treatments
(`r papaja::apa_print(t.test(likable ~ Condition, data = study2_time3, var.equal = T))$statistic`,
and
`r papaja::apa_print(t.test(warmAvg ~ Condition, data = study2_time3, var.equal = T))$statistic`,
respectively).

@@

In a context where advice communicates ego-relevant information (here,
people's attractiveness), we find that the advice people give is
contingent on their incentives. Specifically, when they get rewarded for
being more likeable, they recommend that the advisee bet on a more
favorable rank than when they are incentivized for accuracy.
Importantly, advisees do not discount flattering advice and instead
evaluate people who advise them to bet on a more attractive rank as
warmer and more likeable. These gains to interpersonal perceptions do
not come at the cost of trustworthiness, even when the advice is
substantially inflated (e.g., a recommendation that the advisee is the
most attractive in the group).

@@

We begin by examining whether advisers take the expectations of advisees
into account, even when those expectations are not relevant to the
decision the advisee has to make. In a three-stage experiment, we first
recruit a sample of advisees and ask them to complete a mathematics
quiz. After answering ten multiple-choice questions, participants are
anchored to low or high performance expectations and asked to guess how
many questions they answered correctly. Next, we recruit advisers and
show them either only the true performance of the advisee, or the true
performance and how many questions they guessed they answered correctly.
Advisers are tasked with recommending whether the advisee should compete
against a group of high-performers or a group of low-performers on the
mathematics quiz, where a potential bonus depends on outperforming the
competitor and which group was chosen. Finally, we invite the advisees
back and show them the advice they have received. Advisees then pick
whether to compete against high or low performers, with the outcome
determined by their past performance on the quiz. Our key hypotheses are
that (1) advisees who were anchored to high expectations would be more
likely to receive advice to compete against the high performer group,
and (2) that men are more likely than women to be advised to compete
against the high-performer group when expectations are shown.

@@

As part of our main experiment, participants choose to compete against a
group of previous participants and have their expectations anchored
based on their performance. Thus, we begin by first recruiting a sample
of `r scales::comma(dim(study3_math50)[1])` participants from Prolific
to complete a 10-question multiple choice mathematics quiz. The
questions were taken from a paper-version of the ASVAB standardized
exam, such that answers were not available online. Participants had five
minutes to answer the quiz and were paid 10 cents for each correctly
answered question. We define the top 20 scorers as the "High Performers"
and the bottom 20 scorers as the "Low Performers." On average,
participants answered `r round(mean(study3_math50$Score),2)` questions
correctly, High Performers scored between `r HP_lower` and `r HP_upper`,
and Low Performers scored between `r LP_lower` and `r LP_upper`. In
order to anchor the expectations of participants in our main experiment,
we simulated 1,000 pairings of groups of five participants, with the
5<sup>th</sup> percentile of groups scoring an average of `r p5` and the
95 th percentile scoring an average of `r p95`. We report these averages
to participants in the Low Expectations and High Expectations treatment,
respectively.

@@

We then recruited `r scales::comma(dim(study3_time2_giver)[1])`
participants for Stage 2, placing them in the role of advisers. We began
by informing them of the mathematics quiz that participants in the
preliminary study and Stage 1 had completed, and informed them of the
average score of all participants in the preliminary study. Advisers had
to recommend whether an advisee should compete against the Low
Performers or the High Performers (we used these terms in the survey).
We anticipated that being told to compete against High Performers is
more flattering and hence being told to compete against the Low
Performers would be disappointing if one had expected to do well.
Advisees would earn 30 cents if their score was equal to or higher than
a randomly selected member of the Low Performer group and they chose to
compete against this group, or 50 cents if they performed as well or
better than a randomly selected member of the High Performer group and
they chose to compete against that group. If they scored lower, they
would not receive any bonus.

Advisers were randomly assigned to one of two treatments. In the
"Baseline" treatment, they only observed the score of the advisee on the
mathematics quiz. In the "Expectation" treatment, they observed the
score as well as the advisee's guess for how many questions they
answered correctly. Note that since the outcome is determined only by
the past score on the quiz, the advisee's guess is immaterial to which
group they should compete against. Moreover, in neither treatment did
they receive any demographic information about their advisee.
Participants gave recommendations to 10 advisees, which unbeknownst to
them were five men and five women matched to have identical performance
on the test.[^1] They were informed that if their advice was shown to a
participant who returned for the follow-up survey, they would receive
the identical bonus as that participant. The survey then concluded with
basic demographic questions.

@@

```{r study3genderdiff, fig.cap ="Advisers observed real performance and the participants' estimated performance, but not their gender. As a result of their higher expectations, men (blue) were advised to compete against the high performance group more often than women (red). The grey line shows advice absent expectations, which did not differ by gender."}
library(dplyr)
library(ggplot2)
library(magrittr)

ProbHG <- study3_advice %>%
  group_by(Condition, Performance, Gender) %>%
  dplyr::summarize(avg_hardGroup = mean(Advice))

df_exp <- subset(ProbHG, Condition == "E")
df_none <- subset(ProbHG, Condition == "None")

ggplot(df_exp, aes(x = Performance, y = avg_hardGroup, color = Gender)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),se = FALSE) +
  geom_smooth(data = df_none, aes(x = Performance, y = avg_hardGroup),
              method = "glm", method.args = list(family = "binomial"), color = "grey", se = FALSE) +
  scale_color_manual(values = c("#2A629A",'#FF7F3E', "grey"),
                     labels = c("Female + Expectations", "Male + Expectations", "Expectation Hidden")) +
  scale_y_continuous(name = "P(Compete Against High Performance)",
                     labels = scales::percent,
                     expand = c(0, 0)) +
  scale_x_continuous(name = "Performance",
                     expand = c(0, 0),
                     breaks = seq(0, 10, 2)) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20))
```

@@

<!-- Amanda: The gender discrepancy in advice is more significant when performance is less decisive. When performance is extremely low, there is no chance to win high performers and only a slim chance when competing with low performers. When performance is extremely high, it makes no sense to forgo the high chance to win more and choose a low bonus. However, in the middle range of performance, either option has pros and cons, and neither is absolutely better. In such a situation, advisers would take participants' expectations into account, regarding them as a critical factor. -->

<!-- David: If we plotted on the x-axis performance and on the y-axis the difference -->

<!-- between recommendations for men & women, this would look inverse-U shaped, right? -->

<!-- Might be a nice way to illustrate what you describe. -->

<!-- Amanda: I think our primary purpose here is still showing the existence of gender difference -->

<!-- The current plot shows discrepancy between two lines, allowing us to demonstrate gender gap -->

@@

 men underestimate
their score on a mathematics test less than do women. When these
expectations are displayed to advisers, they are more likely to tell men
to compete against the High Performer group. Notably, this is bad
advice: men whose advisers observed their expectations got worse advice
as a result. Our findings suggest that males with the same scores as
their female counterparts earned less, although the evidence is only
directional. This may be due to males receiving more favorable advice
and being more likely to adopt it, resulting in even worse outcomes.

>>>>

So far, we have shown that displaying expectations influences the advice given. 
Specifically, advisers were more likely to recommend competing against a group 
of high performers to advisees with higher performance expectations compared to 
those without. We now directly test whether this effect is driven by 
interpersonal concerns. In particular, we explore whether more flattering advice
stems from a desire to be liked, and whether advisees, in turn, like advisers 
more when they receive flattering advice.

We conducted a three-stage experiment in which we experimentally
manipulated the incentives for advisers. We began by inviting a group of
advisees to upload photos of themselves ("selfies") and informed them
that they would be rated on their attractiveness. We grouped them with
nine other participants of the same gender and recruited participants of
the opposing gender to rank them from most to least attractive and to
provide advice. Specifically, we asked them to advise the participant
they ranked as the 7th most attractive (4th least attractive) on what
rank they should bet they were ranked by a larger group of raters.
Advisers were randomly assigned to two treatments, receiving a bonus
payment either if the advisee guessed their rank accurately or if the
advisee evaluated the adviser as likeable as measured by an
unincentivized scale response. We hypothesize that advisers who want to
be liked by advisees will recommend that they bet on a lower rank, i.e.
that they are more attractive.

@@

After submitting their ratings, they saw the photo of the participant
they had ranked as the 7th most attractive (i.e., the 4th least
attractive). They were reminded of the rank they had given to that person
and informed that this participant would be invited back and could earn
a \$1 bonus if they guessed their rank correctly. The rank was
determined by the aggregate ratings of all participants who had ranked
this group. Their task was to give advice to this participant about the
rank they should bet on based on having observed all ten selfies and
their own assessment. We randomly assigned participants to one of two
incentivization schemes. In the "Accuracy" treatment, they received a
bonus identical to the advisee: \$1 if they guessed their rank
correctly. In the "Likeability" treatment, we informed them that they
would be rated by the advisee on a 5-point Likert scale for how likeable
they thought they were. Each point on the scale would translate to a
bonus of 20 cents. They then selected a rank that they would recommend
the advisee to bet on.

@@

@@

@@

library(dplyr)
library(ggplot2)
library(reshape2)

study2_time3 <- study2_time3 %>% dplyr::mutate(Guess = as.numeric(Guess), Advice = as.numeric(Advice)) %>%
  dplyr::mutate(HorL = ifelse(Guess == Advice,0, ifelse(Guess > Advice, 1,-1))) %>%
  dplyr::mutate(HorL = as.factor(HorL)) %>% 
  dplyr::mutate(bonus = ifelse(bet == `Final Rank`,1,0)) %>%
  dplyr::mutate(accuracy = `Final Rank`-Advice)

ratings <- study2_time3 %>%
  group_by(HorL) %>%
  summarise(like = round(mean(likable),2), warmth = round(mean(warmAvg),2))

rating_table <- matrix(c(ratings$like,ratings$warmth), nrow = 2, byrow = TRUE, dimnames = list( c("Likable",'Warm'),c("Lower than", "Equal to", "Higher than")))

@@

# Create a data frame similar to your structure
data <- rating_df

# Reshape the data to wide format for the heatmap
heatmap_data <- dcast(data, Condition ~ Advice, value.var = "Freq")

# Melt the data for ggplot
melted_data <- melt(heatmap_data, id.vars = "Condition", variable.name = "Advice", value.name = "Rating")

# Create the heatmap
ggplot(melted_data, aes(x = Advice, y = Condition, fill = Rating)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "grey80", high = "grey30", name = "Rating") +  # Grey gradient from light to dark
  geom_text(aes(label = sprintf("%.2f", Rating)), color = "white", size = 8) +
  theme_minimal() +
  labs(x = "Attractiveness (Advice Relative to Self-Expectations)", y = "Evaluation") +
  theme(
    axis.text.x = element_text(size = 12),  # X-axis text size
          axis.text.y = element_text(size = 12),
          axis.title.x = element_text(size = 12),  # X-axis title size
          axis.title.y = element_text(size = 12))

@@

We were not powered to do a comparison across the two experimental
groups and did not preregister such a difference. Indeed, we find no
difference in likeability and warmth across the two treatments
(`r papaja::apa_print(t.test(likable ~ Condition, data = study2_time3, var.equal = T))$statistic`,
and
`r papaja::apa_print(t.test(warmAvg ~ Condition, data = study2_time3, var.equal = T))$statistic`,
respectively). We also assess the quality of advice by measuring the 
discrepancy between the advised rank and the final rank. Our analysis
revealed no significant difference between the two treatments in terms 
of how helpful the advice was for advisees in making accurate guesses,
`r papaja::apa_print(t.test(accuracy ~ Condition, data = study2_time3, var.equal = T))$statistic`.

@@

In another context where advice communicates ego-relevant information (here,
people's attractiveness), we find that the advice people give is
contingent on their incentives. Specifically, when they get rewarded for
being more likeable, they recommend that the advisee bet on a more
favorable rank than when they are incentivized for accuracy.
Importantly, advisees do not discount flattering advice and instead
evaluate people who advise them to bet on a more attractive rank as
warmer and more likeable. These gains to interpersonal perceptions do
not come at the cost of trustworthiness, even when the advice is
substantially inflated (e.g., a recommendation that the advisee is the
most attractive in the group).

@@

In Study 3, we replicate the main effect in a different ego-relevant domain: 
intelligence. The experimental design is conceptually identical to Study 1, 
but this time we utilize a more direct measure of expectation by asking 
participants to estimate their scores.

This three-stage experiment begins with measuring actual performance. 
Participants are then anchored to low or high performance expectations 
and asked to guess how many questions they answered correctly. As in 
previous studies, we recruit advisers who are presented with either 
just the advisee's true performance or both the true performance and 
the number of questions the advisee believes they answered correctly. 
Advisers are then asked to provide advice on choosing a competitor. 
Finally, we present the advice to the advisees and ask them to make 
their final choice. Our key hypotheses are: (1) advisees anchored to 
high expectations will be more likely to receive advice to compete 
against the high-performing group, and (2) men will be more likely 
than women to be advised to compete against the high-performing group
when expectations are disclosed.

@@

We begin by first recruiting a sample
of `r scales::comma(dim(study3_math50)[1])` participants from Prolific
to complete a 10-question multiple choice mathematics quiz. The
questions were taken from a paper-version of the ASVAB standardized
exam, such that answers were not available online. Participants had five
minutes to answer the quiz and were paid 10 cents for each correctly
answered question. Like before, we define top 20 scores as "High Performers"
and the bottom 20 scorers as the "Low Performers." On average,
participants answered `r round(mean(study3_math50$Score),2)` questions
correctly, High Performers scored between `r HP_lower` and `r HP_upper`,
and Low Performers scored between `r LP_lower` and `r LP_upper`. In
order to anchor the expectations of participants in our main experiment,
we simulated 1,000 pairings of groups of five participants, with the
5<sup>th</sup> percentile of groups scoring an average of `r p5` and the
95 th percentile scoring an average of `r p95`. We report these averages
to participants in the Low Expectations and High Expectations treatment,
respectively.

@@

We then recruited `r scales::comma(dim(study3_time2_giver)[1])`
participants for Stage 2, placing them in the role of advisers. We began
by informing them of the mathematics quiz that participants in the
preliminary study and Stage 1 had completed, and informed them of the
average score of all participants in the preliminary study. Advisers had
to recommend whether an advisee should compete against the Low
Performers or the High Performers (we used these terms in the survey).
We anticipated that being told to compete against High Performers is
more flattering and hence being told to compete against the Low
Performers would be disappointing if one had expected to do well. Advisees
could earn a bonus if their score was equal to or higher than that of 
a randomly selected member from their chosen group. Competing against 
the High Performers was more challenging, resulting in an average bonus
of 50 cents, compared to 30 cents when competing against other groups.

Like before, advisers were randomly assigned to one of two treatments. In the
"Baseline" treatment, they only observed the score of the advisee on the
mathematics quiz. In the "Expectation" treatment, they observed the
score as well as the advisee's guess for how many questions they
answered correctly. We want to emphasize that sice the outcome is determined only by
the past score on the quiz, the advisee's guess is immaterial to which
group they should compete against. Moreover, in neither treatment did
they receive any demographic information about their advisee.
Participants gave recommendations to 10 advisees, which unbeknownst to
them were five men and five women matched to have identical performance
on the test.[^1] They were informed that if their advice was shown to a
participant who returned for the follow-up survey, they would receive
the identical bonus as that participant. The survey then concluded with
basic demographic questions.

@@

```{r study3genderdiff, fig.cap ="Advisers observed real performance and the participants' estimated performance, but not their gender. As a result of their higher expectations, men (blue) were advised to compete against the high performance group more often than women (red). The grey line shows advice absent expectations, which did not differ by gender."}
library(dplyr)
library(ggplot2)

# Calculate averages for the performance, grouped by condition and gender
ProbHG <- study3_advice %>%  
  group_by(Condition, Performance, Gender) %>%  
  summarise(avg_hardGroup = mean(Advice), .groups = 'drop')

# Create a data frame for the NONE condition as the baseline
baseline <- ProbHG %>%  
  filter(Condition == "None") %>%  
  select(Performance, baseline_avg = avg_hardGroup)

# Merge with the original data to get comparison values
comparison_data <- ProbHG %>%  
  filter(Condition != "None") %>%  
  left_join(baseline, by = "Performance") %>%  
  mutate(diff = avg_hardGroup - baseline_avg)  # Calculate the difference

# Create a scatter plot comparing conditions with NONE as baseline
ggplot() +
  # Step lines for the NONE condition
  geom_segment(data = baseline,
               aes(x = Performance, 
                   xend = lead(Performance, 1),  # Create horizontal lines to the next performance point
                   y = baseline_avg, 
                   yend = baseline_avg,
                   linetype = "None"), 
               color = "grey70", size = 0.5, 
               na.rm = TRUE,
               show.legend = TRUE) + 
  geom_point(data = comparison_data %>% filter(Gender == "Male"), 
             aes(x = Performance + 0.5, y = avg_hardGroup, shape = Gender, color = Gender), 
             size = 4) +  # Males with an offset to the left
  geom_point(data = comparison_data %>% filter(Gender == "Female"), 
             aes(x = Performance + 0.5, y = avg_hardGroup, shape = Gender, color = Gender), 
             size = 4) +  # Females with an offset to the right
  scale_shape_manual(values = c(16, 17), labels = c("Female", "Male")) +  # Circle for Females, Triangle for Males
  scale_color_manual(values = c( "#FF7F3E","#2A629A"), 
                     labels = c("Female", "Male")) +  
  scale_y_continuous(name = "P(Compete Against High Performance)",
                     labels = scales::percent,
                     limits = c(0, 1.1),
                     breaks = c(0,1)) +  # Set limits for y-axis
  scale_x_continuous(name = "Performance",
                     limits = c(0, 11),expand = c(0, 0), 
                     breaks = seq(0, 10, 1)) +  
  theme_bw() +  
  theme(panel.grid = element_blank(), 
        panel.border = element_blank(), 
        axis.line = element_line(colour = "grey"), 
        legend.position = "bottom", 
        legend.title = element_blank(), 
        legend.key.size = unit(1.2, "lines"), 
        legend.text = element_text(size = 12), 
        plot.title = element_text(hjust = 0.5, size = 20))

@@

@@

We found supportive evidence for our argument in a different domain. 
Advisors tend to consider advisees' self-expectations and match 
their advice accordingly, aiming to avoid disappointment. Consistent
with the confidence gap narrative, men tend to underestimate their 
mathematics test scores less than women. When these expectations are
displayed to advisors, they are more likely to advise men to compete
against high performers. Notably, this turns out to be poor advice: 
men whose advisors were aware of their expectations received worse 
advice. Our findings suggest that men with the same scores as their
female counterparts ended up earning less, although this result is 
only directional. This discrepancy may be due to men receiving more
favorable advice, being more likely to follow it, and ultimately 
facing worse outcomes.
